---
title: "01-loading_and_normalizingUMCU"
author: "Wenzel Hackeng"
date: "22/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Pre-processing**
First call the required packages
```{r message=FALSE, warning=FALSE}
library("ChAMP")
```

# ChAMP load data 
During filtering, manually check proportion of failed probes to see if there are bad samples. Of note, during load, filtering on quality is not performed, as we will not perform filtering on quality in test sets
```{r message=FALSE, warning=FALSE}
dir.create("01_preprocessing")
filesUMCU<-'./RAW/UMCU/IdatFiles/'
set.seed(11)
myloadUMCU <- champ.load(directory = filesUMCU,
                              method="ChAMP",
                              methValue="B",
                              autoimpute=FALSE, # not default
                              filterDetP=TRUE,
                              ProbeCutoff=0,
                              SampleCutoff=0.1,
                              detPcut=0.01,
                              filterBeads=TRUE,
                              beadCutoff=0.05,
                              filterNoCG=TRUE,
                              filterSNPs=TRUE,
                              population=NULL,
                              filterMultiHit=TRUE,
                              filterXY=TRUE,
                              force=FALSE,
                              arraytype="EPIC")
```
# ChAMP Quality Control 
Check density plots for sample quality. 
```{r message=FALSE, warning=FALSE}

# 1. Quality Control files
dir.create("./01_preprocessing/CHAMP_QCimages")
champ.QC(beta = myloadUMCU$beta,
         pheno=myloadUMCU$pd$Sample_Group,
         mdsPlot=TRUE,
         densityPlot=TRUE,
         dendrogram=TRUE,
         PDFplot=TRUE,
         Rplot=TRUE,
         Feature.sel="None",
         resultsDir = "./01_preprocessing/CHAMP_QCimages/UMCU/")
```

# ChAMP Normalisation
Normalisation, BMIQ is used voor data as Karphathakis' data was BMIQ normalized (BMIQ best method)

```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/Normalisation")
mynormUMCU<-champ.norm(beta=myloadUMCU$beta, rgSet=myloadUMCU$rgSet, mset=myloadUMCU$mset, 
                    resultsDir = "./01_preprocessing/Normalisation/UMCU/", 
                    method="BMIQ", 
                    plotBMIQ=TRUE, 
                    arraytype="EPIC",
                    cores=4)
```

# Save normalized data for later use
```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/PreProcessedDataSets")
dir.create("./01_preprocessing/PreProcessedDataSets/UMCU")
save(myloadUMCU, mynormUMCU, file = './01_preprocessing/PreProcessedDataSets/UMCU/preprocessedUMCU.rData')
```

