---
title: "01-loading_and_normalizingPanNET"
author: "Wenzel Hackeng"
date: "22/10/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Pre-processing**
First call the required packages
```{r message=FALSE, warning=FALSE}
library("GEOquery")
library("ChAMP")
```
# GEO load Pancreatic Neuroendocrine Tumors
Now load the first data (Chan et al. 2018)
```{r eval=FALSE}
dir.create("RAW")

getGEOSuppFiles("GSE117852", makeDirectory = TRUE, baseDir =  './RAW')
untar('./RAW/GSE117852/GSE117852_RAW.tar', exdir = './RAW/GSE117852/IdatFiles')
```
Now manually untar idat files and remove others. Put PD file in IdatFiles folder

# ChAMP load data 
During filtering, manually check proportion of failed probes to see if there are bad samples.

```{r message=FALSE, warning=FALSE}
dir.create("01_preprocessing")
filesGSE117852<-'./RAW/GSE117852/IdatFiles/'
set.seed(11)
myloadGSE117852 <- champ.load(directory = filesGSE117852,
                              method="ChAMP",
                              methValue="B",
                              autoimpute=TRUE,
                              filterDetP=TRUE,
                              ProbeCutoff=0,
                              SampleCutoff=0.1,
                              detPcut=0.01,
                              filterBeads=TRUE,
                              beadCutoff=0.05,
                              filterNoCG=TRUE,
                              filterSNPs=TRUE,
                              population=NULL,
                              filterMultiHit=TRUE,
                              filterXY=TRUE, # not default
                              force=FALSE,
                              arraytype="450K")
```
# ChAMP Quality Control 
Check density plots for sample quality. 
```{r message=FALSE, warning=FALSE}

# 1. Quality Control files
dir.create("./01_preprocessing/CHAMP_QCimages")
champ.QC(beta = myloadGSE117852$beta,
         pheno=myloadGSE117852$pd$Sample_Group,
         mdsPlot=TRUE,
         densityPlot=TRUE,
         dendrogram=TRUE,
         PDFplot=TRUE,
         Rplot=TRUE,
         Feature.sel="None",
         resultsDir = "./01_preprocessing/CHAMP_QCimages/GSE117852/")
```

# ChAMP Normalisation
Normalisation, BMIQ is used voor data as Karphathakis' data was BMIQ normalized (BMIQ best method)

```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/Normalisation")
mynormGSE117852<-champ.norm(beta=myloadGSE117852$beta, rgSet=myloadGSE117852$rgSet, mset=myloadGSE117852$mset, 
                    resultsDir = "./01_preprocessing/Normalisation/GSE117852/", 
                    method="BMIQ", 
                    plotBMIQ=TRUE, 
                    arraytype="450k",
                    cores=4)
```

# Save normalized data for later use
```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/PreProcessedDataSets")
dir.create("./01_preprocessing/PreProcessedDataSets/GSE117852")
save(myloadGSE117852, mynormGSE117852, file = './01_preprocessing/PreProcessedDataSets/GSE117852/preprocessedGSE117852.rData')
```

