---
title: "01-loading_and_normalizingPan_ilealNET"
author: "Wenzel Hackeng"
date: "22/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Pre-processing**
First call the required packages
```{r message=FALSE, warning=FALSE}
library("GEOquery")
library("ChAMP")
```
# GEO load data of gastric, duodenal, pancreatic and small intestinal sporadic and MEN1 associated Neuroendocrine Tumors (Carcinoids)
Now load the first data (Tirosh et al. 2020)
```{r eval=FALSE}
dir.create("RAW")

getGEOSuppFiles("GSE134089", makeDirectory = TRUE, baseDir =  './RAW')
untar('./RAW/GSE134089/GSE134089_RAW.tar', exdir = './RAW/GSE134089/IdatFiles')
```
Now manually untar idat files and remove others. Put PD file in IdatFiles folder

# ChAMP load data 
During filtering, manually check proportion of failed probes to see if there are bad samples.

```{r message=FALSE, warning=FALSE}
dir.create("01_preprocessing")
filesGSE134089<-'./RAW/GSE134089/IdatFiles/'
set.seed(11)
myloadGSE134089 <- champ.load(directory = filesGSE134089,
                              method="ChAMP",
                              methValue="B",
                              autoimpute=FALSE, # not default
                              filterDetP=TRUE,
                              ProbeCutoff=0,
                              SampleCutoff=0.1,
                              detPcut=0.01,
                              filterBeads=TRUE,
                              beadCutoff=0.05,
                              filterNoCG=TRUE,
                              filterSNPs=TRUE,
                              population=NULL,
                              filterMultiHit=TRUE,
                              filterXY=TRUE,
                              force=FALSE,
                              arraytype="EPIC")
```

# Free up RAM
```{r Free space, message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/PreProcessedDataSets")
dir.create("./01_preprocessing/PreProcessedDataSets/GSE134089")
save(myloadGSE134089, file = './01_preprocessing/PreProcessedDataSets/GSE134089/myloadGSE134089.rData')

#Free RAM
rm(list=ls())
gc() 

#reload
library("ChAMP")
library("readr")
load('./01_preprocessing/PreProcessedDataSets/GSE134089/myloadGSE134089.rData')

```

# ChAMP Quality Control 
Check density plots for sample quality. 
```{r message=FALSE, warning=FALSE}
# 1. Quality Control files
dir.create("./01_preprocessing/CHAMP_QCimages")
champ.QC(beta = myloadGSE134089$beta,
         pheno=myloadGSE134089$pd$Sample_Group,
         mdsPlot=TRUE,
         densityPlot=TRUE,
         dendrogram=TRUE,
         PDFplot=TRUE,
         Rplot=TRUE,
         Feature.sel="None",
         resultsDir = "./01_preprocessing/CHAMP_QCimages/GSE134089/")
```

# ChAMP Normalisation
Normalisation, BMIQ is recommended 
```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/Normalisation")
mynormGSE134089<-champ.norm(beta=myloadGSE134089$beta, rgSet=myloadGSE134089$rgSet, mset=myloadGSE134089$mset, 
                    resultsDir = "./01_preprocessing/Normalisation/GSE134089/", 
                    method="BMIQ", 
                    plotBMIQ=TRUE, 
                    arraytype="EPIC",
                    cores=4)
```

```{r}
PD_GSE134089 <- read_csv("./PD_FILES/PD_GSE134089.csv", 
    skip = 7)
PD_GSE134089<-PD_GSE134089[!(PD_GSE134089$Sample_Name=="B08_tumor"),] # low quality sample removed in preprocessing
PD_GSE134089<-PD_GSE134089[!(PD_GSE134089$Sample_Name=="F08_tumor"),] # low quality sample removed in preprocessing


PD_GSE134089cases<-subset(PD_GSE134089, Sample_Group == "IlealNET" | Sample_Group == "PanNET"| Sample_Group == "UPNET")
PD_GSE134089otherGI<-subset(PD_GSE134089, Sample_Group == "DuoNET" | Sample_Group == "GasNET")


mynormGSE134089cases <- as.data.frame(mynormGSE134089[,colnames(mynormGSE134089) %in% PD_GSE134089cases$Sample_Name])
mynormGSE134089otherGI <- as.data.frame(mynormGSE134089[,colnames(mynormGSE134089) %in% PD_GSE134089otherGI$Sample_Name])

```


# Save normalized data for later use
```{r message=FALSE, warning=FALSE}
save(mynormGSE134089cases, mynormGSE134089otherGI, PD_GSE134089cases, PD_GSE134089otherGI, file = './01_preprocessing/PreProcessedDataSets/GSE134089/preprocessedGSE134089.rData')

```

