---
title: "01-loading_and_normalizingPulmNET"
author: "Wenzel Hackeng"
date: "22/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Pre-processing**
First call the required packages
```{r message=FALSE, warning=FALSE}
library("GEOquery")
library("ChAMP")
```
# GEO load Pulmonary Neuroendocrine Tumors (Carcinoids)
Now load the first data (Laddha et al. 2019)
```{r eval=FALSE}
dir.create("RAW")

getGEOSuppFiles("GSE118133", makeDirectory = TRUE, baseDir =  './RAW')
untar('./RAW/GSE118133/GSE118133_RAW.tar', exdir = './RAW/GSE118133/IdatFiles')
```
Now manually untar idat files and remove others. Put PD file in IdatFiles folder

# ChAMP load data 
During filtering, manually check proportion of failed probes to see if there are bad samples.

```{r message=FALSE, warning=FALSE}
dir.create("01_preprocessing")
filesGSE118133<-'./RAW/GSE118133/IdatFiles/'
set.seed(11)
myloadGSE118133 <- champ.load(directory = filesGSE118133,
                              method="ChAMP",
                              methValue="B",
                              autoimpute=TRUE,
                              filterDetP=TRUE,
                              ProbeCutoff=0,
                              SampleCutoff=0.1,
                              detPcut=0.01,
                              filterBeads=TRUE,
                              beadCutoff=0.05,
                              filterNoCG=TRUE,
                              filterSNPs=TRUE,
                              population=NULL,
                              filterMultiHit=TRUE,
                              filterXY=TRUE,
                              force=FALSE,
                              arraytype="450K")
```
# ChAMP Quality Control 
Check density plots for sample quality. 
```{r message=FALSE, warning=FALSE}

# 1. Quality Control files
dir.create("./01_preprocessing/CHAMP_QCimages")
champ.QC(beta = myloadGSE118133$beta,
         pheno=myloadGSE118133$pd$Sample_Group,
         mdsPlot=TRUE,
         densityPlot=TRUE,
         dendrogram=TRUE,
         PDFplot=TRUE,
         Rplot=TRUE,
         Feature.sel="None",
         resultsDir = "./01_preprocessing/CHAMP_QCimages/GSE118133/")
```

# ChAMP Normalisation
Normalisation, BMIQ is used voor data as Karphathakis' data was BMIQ normalized (BMIQ best method)

```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/Normalisation")
mynormGSE118133<-champ.norm(beta=myloadGSE118133$beta, rgSet=myloadGSE118133$rgSet, mset=myloadGSE118133$mset, 
                    resultsDir = "./01_preprocessing/Normalisation/GSE118133/", 
                    method="BMIQ", 
                    plotBMIQ=TRUE, 
                    arraytype="450k",
                    cores=4)
```

# Save normalized data for later use

```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/PreProcessedDataSets")
dir.create("./01_preprocessing/PreProcessedDataSets/GSE118133")
save(myloadGSE118133, mynormGSE118133, file = './01_preprocessing/PreProcessedDataSets/GSE118133/preprocessedGSE118133.rData')

```

