---
title: "01-loading_and_normalizingIlealNET"
author: "Wenzel Hackeng"
date: "22/10/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Pre-processing**
First call the required packages
```{r message=FALSE, warning=FALSE}
library("GEOquery")
library("ChAMP")
library("readxl")
library("readr")
library("plyr")
```
# GEO load Small Intestinal Neuroendocrine tumors
Download Phenotype data and link to case information. Put this file and the PD file in TextFiles folder (GEO data and the supplementary files from Karpathakis et al. 2016 were linked based on age, sex, FF-FFPE, Ki-67 count, Tumor-normal in this excel file). Now load all seperate GSM.soft data (Karpathakis et al. 2016). 
 
```{r eval=FALSE}
dir.create("RAW")
gset <- getGEO("GSE73832", GSEMatrix =TRUE, getGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL13534", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]
GSE73832_phenodata<- as.data.frame(pData(gset))

library(readxl)
Karpathakis_Linking_Cases <- read_excel("./RAW/GSE73832/TextFiles/Karpathakis_Linking_Cases.xlsx", sheet = "Combined GEO and Supplementary")
GSMs<-Karpathakis_Linking_Cases$GSM_num

dir.create("./RAW/GSE73832/GSMFiles")
for(i in GSMs){
getGEOfile(i, destdir = './RAW/GSE73832/GSMFiles', AnnotGPL = FALSE, amount = c("data"))
}
```

# Data wrangling
Of note, these data are already BMIQ normalized, but not filtered. 
```{r message=FALSE, warning=FALSE}
Karpathakis_Linking_Cases <- read_excel("./RAW/GSE73832/TextFiles/Karpathakis_Linking_Cases.xlsx", sheet = "Combined GEO and Supplementary")
GSMs<-Karpathakis_Linking_Cases$GSM_num

GSMdata <-list()
for (i in GSMs){
temp <- as.data.frame(read_table2(paste0("./RAW/GSE73832/GSMFiles/", i, ".soft", sep = ""), skip = 5))
temp$Pval<-NULL
colnames(temp)[3]<-"Pval"
GSMdata[[i]]<-temp
}


# select ileal cases for further analysis
IlealGSE73832<- subset(Karpathakis_Linking_Cases, 
                       Sample_Group == "Primary Ileal" &
                         Sample_Type == "FFPE")
SIotherGSE73832<- subset(Karpathakis_Linking_Cases, 
                       Sample_Group %in% c(
                      "LiverMet SI NOS","Met SI NOS","OmenMet SI NOS","Primary SI_NOS")
                         & Sample_Type %in% c("FFPE"))
GIotherGSE73832<- subset(Karpathakis_Linking_Cases, 
                       Sample_Group %in% c(
                      "OmenMet Colon","Primary Appendix","Primary Colon","Primary DuodJej", "Primary Gastric")
                         & Sample_Type %in% c("FFPE"))


GSMsileal<-IlealGSE73832$GSM_num
GSMsSIother<-SIotherGSE73832$GSM_num
GSMsGIother<-GIotherGSE73832$GSM_num

table(as.factor(Karpathakis_Linking_Cases$Sample_Group), as.factor(Karpathakis_Linking_Cases$Sample_Type))
```

# Data  filtering
import newly made pd and filter. NB not filtered for beadcount! this information not provided the the GEO raw data. Determine % of probes with detection P-values >0.01. Select beta values from expression data.
```{r message=FALSE, warning=FALSE}
# function to check the fraction of failed probes
FailProb<-function(dfinlist){
  percFailProb<-nrow(dfinlist[dfinlist$Pval>0.01, ])/nrow(dfinlist)
}

# function to remove the failed probes 
FilterOnSig<-function(dfinlist){
  FilteredDF<-dfinlist[dfinlist$Pval<=0.01,]
}

# function to prep dataframes for merging betas and further analysus
PrepGSMs<-function(dfinlist){
  dfinlist["Pval"] <- NULL; dfinlist 
}

# Fraction of failed probes
failedprobes<-lapply(GSMdata, FailProb)
failedprobesileal<-as.data.frame(unlist(failedprobes[GSMsileal]))
colnames(failedprobesileal)[1]<-"Failed CpG Fraction Ileal NETs"
failedprobesileal

# Fraction of failed probes other
failedprobesSIother<-as.data.frame(unlist(failedprobes[GSMsSIother]))
colnames(failedprobesSIother)[1]<-"Failed CpG Fraction other small intestinal"
failedprobesSIother

# Fraction of failed probes other
failedprobesGIother<-as.data.frame(unlist(failedprobes[GSMsGIother]))
colnames(failedprobesGIother)[1]<-"Failed CpG Fraction other gastro-intestinal"
failedprobesGIother

# removing failed probes and changing beta names
FilteredList<-lapply(GSMdata, FilterOnSig)
for (i in 1:length(GSMdata)) {
  colnames(GSMdata[[i]])[2] <- names(GSMdata)[i]
 } 

# Prepping and merging betavalues
FilteredListbetas<-lapply(FilteredList, PrepGSMs)
for (i in 1:length(FilteredListbetas)) {
  colnames(FilteredListbetas[[i]])[2] <- names(FilteredListbetas)[i]
 } 

# Ileal cases
GSE73832ilealbetas<-join_all(FilteredListbetas[GSMsileal], by = "ID_REF", type = "inner", match = "all")
GSE73832ilealbetas<-GSE73832ilealbetas[complete.cases(GSE73832ilealbetas), ]
row.names(GSE73832ilealbetas)<-GSE73832ilealbetas[,1]
GSE73832ilealbetas[,1]<-NULL

# Other SI cases
GSE73832SIotherbetas<-join_all(FilteredListbetas[GSMsSIother], by = "ID_REF", type = "inner", match = "all")
GSE73832SIotherbetas<-GSE73832SIotherbetas[complete.cases(GSE73832SIotherbetas), ]
row.names(GSE73832SIotherbetas)<-GSE73832SIotherbetas[,1]
GSE73832SIotherbetas[,1]<-NULL

# Other GI cases
GSE73832GIotherbetas<-join_all(FilteredListbetas[GSMsGIother], by = "ID_REF", type = "inner", match = "all")
GSE73832GIotherbetas<-GSE73832GIotherbetas[complete.cases(GSE73832GIotherbetas), ]
row.names(GSE73832GIotherbetas)<-GSE73832GIotherbetas[,1]
GSE73832GIotherbetas[,1]<-NULL

#  ChAMP Filtering for ileal cases
PD_GSE73832<-read.csv("./RAW/GSE73832/TextFiles/PD_GSE73832.csv", header=TRUE)
set.seed(11)
myloadilealGSE73832<-champ.filter(beta=GSE73832ilealbetas,
                      pd=PD_GSE73832,
                      intensity=NULL,
                      Meth=NULL,
                      UnMeth=NULL,
                      detP=NULL,
                      beadcount=NULL,
                      autoimpute=TRUE,
                      filterDetP=TRUE,
                      ProbeCutoff=0,
                      SampleCutoff=0.1,
                      detPcut=0.01,
                      filterBeads=FALSE,
                      beadCutoff=0.05,
                      filterNoCG = TRUE,
                      filterSNPs = TRUE,
                      population = NULL,
                      filterMultiHit = TRUE,
                      filterXY = TRUE,
                      fixOutlier = FALSE,
                      arraytype = "450K")

mynormilealGSE73832<-myloadilealGSE73832[[1]]

#  ChAMP Filtering for other small intestinal cases
PD_SIotherGSE73832<-read.csv("./RAW/GSE73832/TextFiles/PD_SIotherGSE73832.csv", header=TRUE)
myloadSIotherGSE73832<-champ.filter(beta=GSE73832SIotherbetas,
                      pd=PD_SIotherGSE73832,
                      intensity=NULL,
                      Meth=NULL,
                      UnMeth=NULL,
                      detP=NULL,
                      beadcount=NULL,
                      autoimpute=FALSE,
                      filterDetP=FALSE,
                      ProbeCutoff=0,
                      SampleCutoff=0.1,
                      detPcut=0.01,
                      filterBeads=FALSE,
                      beadCutoff=0.05,
                      filterNoCG = TRUE,
                      filterSNPs = TRUE,
                      population = NULL,
                      filterMultiHit = TRUE,
                      filterXY = TRUE,
                      fixOutlier = FALSE,
                      arraytype = "450K")

mynormSIotherGSE73832<-myloadSIotherGSE73832[[1]]

#  ChAMP Filtering for other gastro intestinal cases
PD_GIotherGSE73832<-read.csv("./RAW/GSE73832/TextFiles/PD_GIotherGSE73832.csv", header=TRUE)
myloadGIotherGSE73832<-champ.filter(beta=GSE73832GIotherbetas,
                      pd=PD_GIotherGSE73832,
                      intensity=NULL,
                      Meth=NULL,
                      UnMeth=NULL,
                      detP=NULL,
                      beadcount=NULL,
                      autoimpute=FALSE,
                      filterDetP=FALSE,
                      ProbeCutoff=0,
                      SampleCutoff=0.1,
                      detPcut=0.01,
                      filterBeads=FALSE,
                      beadCutoff=0.05,
                      filterNoCG = TRUE,
                      filterSNPs = TRUE,
                      population = NULL,
                      filterMultiHit = TRUE,
                      filterXY = TRUE,
                      fixOutlier = FALSE,
                      arraytype = "450K")

mynormGIotherGSE73832<-myloadGIotherGSE73832[[1]]

```

# ChAMP Quality Control 
NB. Samples were already normalized.  
```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/CHAMP_QCimages")
champ.QC(beta = myloadilealGSE73832$beta,
         pheno=myloadilealGSE73832$pd$Sample_Group,
         mdsPlot=FALSE, # must be on MethylSet', 'RGChannelSet', or 'matrix'
         densityPlot=FALSE,# must be on MethylSet', 'RGChannelSet', or 'matrix'
         dendrogram=TRUE,
         PDFplot=TRUE,
         Rplot=TRUE,
         Feature.sel="None",
         resultsDir = "./01_preprocessing/CHAMP_QCimages/IlealGSE73832/")

champ.QC(beta = myloadSIotherGSE73832$beta,
         pheno=myloadSIotherGSE73832$pd$Sample_Group,
         mdsPlot=FALSE, # must be on MethylSet', 'RGChannelSet', or 'matrix'
         densityPlot=FALSE,# must be on MethylSet', 'RGChannelSet', or 'matrix'
         dendrogram=TRUE,
         PDFplot=TRUE,
         Rplot=TRUE,
         Feature.sel="None",
         resultsDir = "./01_preprocessing/CHAMP_QCimages/SIotherGSE73832/")

champ.QC(beta = myloadGIotherGSE73832$beta,
         pheno=myloadGIotherGSE73832$pd$Sample_Group,
         mdsPlot=FALSE, # must be on MethylSet', 'RGChannelSet', or 'matrix'
         densityPlot=FALSE,# must be on MethylSet', 'RGChannelSet', or 'matrix'
         dendrogram=TRUE,
         PDFplot=TRUE,
         Rplot=TRUE,
         Feature.sel="None",
         resultsDir = "./01_preprocessing/CHAMP_QCimages/GIotherGSE73832/")
```

# Save normalized data for later use
```{r message=FALSE, warning=FALSE}
dir.create("./01_preprocessing/PreProcessedDataSets")
dir.create("./01_preprocessing/PreProcessedDataSets/IlealGSE73832")
dir.create("./01_preprocessing/PreProcessedDataSets/SIotherGSE73832")
dir.create("./01_preprocessing/PreProcessedDataSets/GIotherGSE73832")
save(myloadilealGSE73832, mynormilealGSE73832, file = './01_preprocessing/PreProcessedDataSets/IlealGSE73832/preprocessedIlealGSE73832.rData')
save(myloadSIotherGSE73832, mynormSIotherGSE73832, file = './01_preprocessing/PreProcessedDataSets/SIotherGSE73832/preprocessedSIotherGSE73832.rData')
save(myloadGIotherGSE73832, mynormGIotherGSE73832, file = './01_preprocessing/PreProcessedDataSets/GIotherGSE73832/preprocessedGIotherGSE73832.rData')

```

