---
title: "02-merge_and_peek"
author: "Wenzel Hackeng"
date: "22/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **2-d visualisation**
First call the required packages and functions
```{r message=FALSE, warning=FALSE}
library('Rtsne')
library('ChAMP')
library('ggplot2')
library('ggrepel')
library('ggsci')
library('caret')
library('matrixStats')
library('pheatmap')
library('cowplot')
```

# Load preprocessed data
```{r message=FALSE, warning=FALSE}
load('./01_preprocessing/PreProcessedDataSets/UMCU/preprocessedUMCU.rData')
load('./01_preprocessing/PreProcessedDataSets/GSE117852/preprocessedGSE117852.rData')
load('./01_preprocessing/PreProcessedDataSets/GSE118133/preprocessedGSE118133.rData')
load('./01_preprocessing/PreProcessedDataSets/IlealGSE73832/preprocessedIlealGSE73832.rData')
PD_combined<-read.csv("./PD_FILES/PD_combined.csv", header=TRUE)
```


```{r message=FALSE, warning=FALSE}
Seed<-11
Cluster<-"euclidean" #"correlation","euclidean"
newdir<-"02_merge_and_peek"
ifelse(!dir.exists(file.path(getwd(), newdir)), dir.create(file.path(getwd(), newdir)), FALSE)
```

# Data merge and save for further processing
Overlapping probes are selected by merging data. 
```{r message=FALSE, warning=FALSE}
df1<- as.data.frame(mynormGSE117852)
df2<- as.data.frame(mynormGSE118133)
df3<- as.data.frame(mynormilealGSE73832)
df4<- as.data.frame(mynormUMCU)

premerge<- merge(df3, df1, by.x="row.names", by.y="row.names")
merged<- merge(premerge, df2, by.x="Row.names", by.y="row.names")
rownames(merged) <- merged[,1]
merged[,1] <- NULL


dir.create("./01_preprocessing/PreProcessedDataSets/GEOMERGED")
save(merged, file = './01_preprocessing/PreProcessedDataSets/GEOMERGED/GEOMERGED.rData')
```

# Select most variable positions (MVP)
Triple downsampling cases until groups are equal. 5000 probes with the highest mean absolute standard deviation
```{r message=FALSE, warning=FALSE}
# add class information
tmerged<-as.data.frame(t(merged))
annoGEO<- merge(PD_combined[,c(1,3)], tmerged, by.x="Sample_Name", by.y="row.names", sort = FALSE)
annoGEO$Sample_Group <- factor(annoGEO$Sample_Group, levels = unique(annoGEO$Sample_Group))

# downsample three times
set.seed(Seed)
downsampnames<-list()
for(i in 1:3){
  downsampnames[[i]]<-downSample(annoGEO[,1],annoGEO$Sample_Group)
}
# merge with beta values and determine mean absolute standard deviation
downsamples<-lapply(downsampnames,function(x,y){merge(x, y, by.x=names(x)[1], by.y="row.names")},tmerged)
downsamples<-lapply(downsamples, "[", -c(1, 2))
downsamples<-lapply(downsamples,function(x){t(x)})

a<-downsamples[[1]]
b<-downsamples[[2]]
c<-downsamples[[3]]

amads=apply(a,1,mad)
bmads=apply(b,1,mad)
cmads=apply(c,1,mad)

amergedmad<-as.data.frame(a[rev(order(amads)),])
bmergedmad<-as.data.frame(b[rev(order(bmads)),])
cmergedmad<-as.data.frame(c[rev(order(cmads)),])

amergedmad$ordera<-1:nrow(amergedmad)
bmergedmad$orderb<-1:nrow(bmergedmad)
cmergedmad$orderc<-1:nrow(cmergedmad)

# selecting probes by the minimal rank of the mean absolute standard deviation over all downsamples. 

MVprobes<-merge(amergedmad, bmergedmad, by.x= "row.names", by.y= "row.names")
MVprobes<-merge(MVprobes, cmergedmad, by.x= "Row.names", by.y= "row.names")
rownames(MVprobes) <- MVprobes[,1]
MVprobes[,1] <-NULL
MVprobes<-MVprobes[,c("ordera","orderb","orderc")]

MVprobes$min <- rowMins(as.matrix(MVprobes))
MVprobes$max <- rowMaxs(as.matrix(MVprobes))

MVprobes <-MVprobes[order(MVprobes$max, decreasing = FALSE),]
bestprobesMVP<-MVprobes[1:5000,c("min", "max")]

# merge with beta valus for further analysis
mergedmad<-merge(bestprobesMVP, merged, by.x= "row.names", by.y= "row.names")
rownames(mergedmad) <- mergedmad[,1]
mergedmad[,1:3] <-NULL

```

# T-sne visualization on complete training cohort
without PCA. Of note, Lung carcinoid type 2 carcinoids show aberrant clustering in T-sne and unsupervised hierarchical clusterig
```{r message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5}

tmergedmad<-as.data.frame(t(mergedmad))
annodatamerged<- merge(tmergedmad, PD_combined, by.x="row.names", by.y="Sample_Name", sort = FALSE)
annodatamerged$Sample_Group <- factor(annodatamerged$Sample_Group, levels = unique(annodatamerged$Sample_Group))
annodatamerged$Sample_Group<-factor(annodatamerged$Sample_Group, levels = c("PanNET", "PulmNET","IlealNET"))

## TSNE with label
set.seed(Seed) # for reproducibility
tsne_out <- Rtsne(tmergedmad,pca=FALSE, 
                  dims = 2, 
                  theta=0, 
                  perplexity = 16, 
                  verbose=F, 
                  max_iter = 5000)

tsne_plot <- data.frame(x = tsne_out$Y[,1], y = tsne_out$Y[,2], col = annodatamerged$Sample_Group, lab = annodatamerged$Row.names)
ggplot(tsne_plot)+ 
  geom_point(aes(x=x, y=y, color=col, size=30))+
  geom_text_repel(aes(x=x, y=y, label=lab), size=3, family="Arial")+
  labs(x = "t-SNE 1", y = "t-SNE 2", col = "Origins")+
  scale_color_npg(labels = c("Chan et al. Pancreas NET", "Laddha et al. Lung NET", "Karpathakis et al. Ileal NET"))+
  theme(
    panel.border = element_blank(),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "grey"),
    legend.key=element_blank())+
  guides(colour = guide_legend(override.aes = list(size=5)))

ggsave("tSNEtraining.eps", path = "./02_merge_and_peek/", width = 10,
  height = 6,
  units = "in")

```


# Unsupervised Hierarchical Clustering complete training cohort 
```{r message=FALSE, warning=FALSE}
Var1        <- c("#DC0000B2", "#4DBBD5B2","purple")
names(Var1) <- c("PanNET","PulmNET","IlealNET")
anno_colors <- list(Var1 = Var1)

pd<-as.data.frame(annodatamerged$Sample_Group[1:69])
row.names(pd)<-annodatamerged[1:69,1]
names(pd)[1]<-"Type"
pd[,1]<-as.character(pd[,1])
set.seed(Seed)
heat<-pheatmap(mergedmad[1:5000,1:69], kmeans_k = NA, breaks = NA, border_color= "grey60",cellwidth = 5, cellheight = 0.05, scale = "none", cluster_rows = TRUE,
         cluster_cols = TRUE, clustering_distance_rows = Cluster,
         clustering_distance_cols = Cluster, clustering_method = "complete",
         cutree_rows = 1, cutree_cols = 1,
         legend_labels = NA, annotation_row = NA, annotation.col = pd,
         annotation = pd, annotation_colors = anno_colors, annotation_legend = TRUE,
         annotation_names_row = TRUE, annotation_names_col = TRUE,
         drop_levels = TRUE, show_rownames = F, show_colnames = T, main = NA, fontsize = 4,
         fontsize_row = 4, fontsize_col = 4,
         display_numbers = F, number_format = "%.2f", number_color = "grey30",
         fontsize_number = 0.8 * fontsize, gaps_row = NULL, gaps_col = NULL,
         labels_row = NULL, labels_col = NULL, filename = NA, width = NA,
         height = NA, silent = FALSE)

save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   pdf(filename, width=width, height=height)
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}
save_pheatmap_pdf(heat, "./02_merge_and_peek/SupFigclustering.pdf")

```

# Merge with UMCU cases and add origin information
Load combined PD files. Merge UMCU cases with the MVPs as selected in the merged online cohorts. Combine origin information of cases  with the most variable positions
```{r message=FALSE, warning=FALSE}
mergedmadumcu<-merge(mergedmad, df4, by.x="row.names", by.y="row.names")
rownames(mergedmadumcu) <- mergedmadumcu[,1]
mergedmadumcu[,1] <- NULL
nrow(mergedmadumcu)

tmergedmadumcu<-as.data.frame(t(mergedmadumcu))
annodatamergedumcu<- merge(tmergedmadumcu, PD_combined, by.x="row.names", by.y="Sample_Name", sort = FALSE)
annodatamergedumcu$Sample_Group <- factor(annodatamergedumcu$Sample_Group, levels = unique(annodatamergedumcu$Sample_Group))

annodatamergedumcu$Sample_Group<-factor(annodatamergedumcu$Sample_Group, levels = c( "PanNET","PulmNET", "IlealNET", "CasePan", "CaseLung", "CaseUP", "CaseIleum"))

```

# T-sne visualization on complete training cohort and UMCU test cohort
without PCA without annotations
```{r message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5}
## TSNE with label
set.seed(Seed) # for reproducibility
tsne_out <- Rtsne(tmergedmadumcu,pca=FALSE, 
                  dims = 2, 
                  theta=0, 
                  perplexity = 16, 
                  verbose=F, 
                  max_iter = 5000)

tsne_plot <- data.frame(x = tsne_out$Y[,1], y = tsne_out$Y[,2], col = annodatamergedumcu$Sample_Group, lab = annodatamergedumcu$Row.names)
ggplot(tsne_plot)+ 
  geom_point(aes(x=x, y=y, color=col, size=30))+
  #geom_text_repel(aes(x=x, y=y, label=lab), size=3, family="Arial")+
  labs(x = "t-SNE 1", y = "t-SNE 2", col = "Origins")+
  scale_color_npg(labels = c("Chan et al. Pancreas NET", "Laddha et al. Lung NET","Karpathakis et al. Ileal NET", "UMCU Pancreas NET", "UMCU Lung NET", "UMCU Unknown Primary", "UMCU Ileal NET"))+
  theme(
    panel.border = element_blank(),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "grey"),
    legend.key=element_blank())+
  guides(colour = guide_legend(override.aes = list(size=5)))

ggsave("tSNEfig1.eps", path = "./02_merge_and_peek/", width = 7.5,
  height = 4,
  units = "in")

```


# T-sne visualization on complete training cohort and UMCU test cohort
without PCA with annotations
```{r message=FALSE, warning=FALSE, fig.width=9, fig.height=4.5}
## TSNE with label
set.seed(Seed) # for reproducibility
tsne_out <- Rtsne(tmergedmadumcu,pca=FALSE, 
                  dims = 2, 
                  theta=0, 
                  perplexity = 16, 
                  verbose=F, 
                  max_iter = 5000)

tsne_plot <- data.frame(x = tsne_out$Y[,1], y = tsne_out$Y[,2], col = annodatamergedumcu$Sample_Group, lab = annodatamergedumcu$Row.names)
ggplot(tsne_plot)+ 
  geom_point(aes(x=x, y=y, color=col, size=30))+
  geom_text_repel(aes(x=x, y=y, label=lab), size=3, family="Arial")+
  labs(x = "t-SNE 1", y = "t-SNE 2", col = "Origins")+
  scale_color_npg(labels = c("Chan et al. Pancreas NET", "Laddha et al. Lung NET","Karpathakis et al. Ileal NET", "UMCU Pancreas NET", "UMCU Lung NET", "UMCU Unknown Primary", "UMCU Ileal NET"))+
  theme(
    panel.border = element_blank(),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "grey"),
    legend.key=element_blank())+
  guides(colour = guide_legend(override.aes = list(size=5)))

ggsave("tSNEfig1_anno.eps", path = "./02_merge_and_peek/", width = 10,
  height = 6,
  units = "in")

```
