---
title: "03-Results3x3fold-nestedCV"
author: "Wenzel Hackeng"
date: "22/10/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First call the required packages 
```{r message=FALSE, warning=FALSE}
library('dplyr')
library('tidyr')
library('ggplot2')
library('cowplot')
```

# Load data
```{r message=FALSE, warning=FALSE}
load("./03_nestedCV/nestedCVREPsseed11rep3tr500llambda.1sea0nf32020-10-22.rData")
```

# Settings of the run above
```{r message=FALSE, warning=FALSE}
# general choices
Seed<-11 # for repeatability
repeats<-3 # repeats of the whole proces of, split, CV and nested CV
# Random forest indicators
Trees<-500
# Ridge Regression indicators
Lambda<-"lambda.1se" # or lambda.min
Alpha <- 0 # 0 = Ridge regression, 1 = Lasso
Nfolds <- 3 #fold nested CV
```

# Results for Accuracy
Boxplot for accuracy based on 9 values per group (3 fold CV with 3 repeats). 
Of note, multiclass ROC sometimes gives an AUC of 1, even though the accuracy is <1. This is because multiclass ROC as defined by Hand and Till compared 1-vs-1 class instead of 1-vs-the rest. 
```{r message=FALSE, warning=FALSE}
# Results
results <- matrix(nrow=3*repeats, ncol=21)
results[,1]<-rep(1:repeats, rep=repeats, each=3)
results[,2]<-Seed
results[,3]<-Trees
results[,4]<-Lambda
results[,5]<-Alpha
results[,6]<-Nfolds

for (r in 1:repeats)
  for (i in 1:3){
  results[((r-1)*3+i),7]<-1-REPs[[r]][[i]][["RF"]][["OOBerror"]]
  results[((r-1)*3+i),8]<-REPs[[r]][[i]][["predictRFtest"]][["Accuracy"]]
  results[((r-1)*3+i),9]<-REPs[[r]][[i]][["predictLRtest"]][["Accuracy"]]
  results[((r-1)*3+i),10]<-REPs[[r]][[i]][["predictRFumcuval"]][["Accuracy"]]
  results[((r-1)*3+i),11]<-REPs[[r]][[i]][["predictLRumcuval"]][["Accuracy"]]
  results[((r-1)*3+i),12]<-REPs[[r]][[i]][["RFtrainROC"]][["auc"]]
  results[((r-1)*3+i),13]<-REPs[[r]][[i]][["predictRFtestROC"]][["auc"]]
  results[((r-1)*3+i),14]<-REPs[[r]][[i]][["predictLRtestROC"]][["auc"]]
  results[((r-1)*3+i),15]<-REPs[[r]][[i]][["predictRFumcuvalROC"]][["auc"]]
  results[((r-1)*3+i),16]<-REPs[[r]][[i]][["predictLRumcuvalROC"]][["auc"]]
  results[((r-1)*3+i),17]<-REPs[[r]][[i]][["RFtrainlogloss"]]
  results[((r-1)*3+i),18]<-REPs[[r]][[i]][["predictRFtestlogloss"]]
  results[((r-1)*3+i),19]<-REPs[[r]][[i]][["predictLRtestlogloss"]]
  results[((r-1)*3+i),20]<-REPs[[r]][[i]][["predictRFumcuvallogloss"]]
  results[((r-1)*3+i),21]<-REPs[[r]][[i]][["predictLRumcuvallogloss"]]
}

results<-as.data.frame(results)
colnames(results)<-c("Repeats", "Seed","RFtrees", "Lambda", 
                     "Alpha", "Nfolds", "Train_OOB_RFaccuracy", "Test_RFaccuracy", "Test_LRaccuracy", "UMCUval_RFaccuracy", "UMCUval_LRaccuracy", "TrainRFAUC", "Test_RFAUC", "Test_LRAUC", "UMCUval_RFAUC", "UMCUval_LRAUC", "TrainRFlogloss", "Test_RFlogloss", "Test_LRlogloss", "UMCUval_RFlogloss", "UMCUval_LRlogloss")



newdir<-"03_nestedCV"
ifelse(!dir.exists(file.path(getwd(), newdir)), dir.create(file.path(getwd(), newdir)), FALSE)
myfile <- file.path(paste0("./03_nestedCV/nestedCVresults","seed",Seed,"rep",repeats,"tr",Trees,"l",Lambda,"a",Alpha,"nf",Nfolds,".rData"))
save(results, file = myfile)

#make long data
longresultsaccuracy<-gather(results, Groups, Accuracy, Train_OOB_RFaccuracy, Test_RFaccuracy, Test_LRaccuracy, UMCUval_RFaccuracy, UMCUval_LRaccuracy, factor_key=TRUE)
longresultsaccuracy$Accuracy<-as.numeric(longresultsaccuracy$Accuracy)
longresultsaccuracy$Accuracy<-longresultsaccuracy$Accuracy*100

longresultsAUC<-gather(results, Groups, AUC, TrainRFAUC, Test_RFAUC, Test_LRAUC, UMCUval_RFAUC, UMCUval_LRAUC, factor_key=TRUE)
longresultsAUC$AUC<-as.numeric(longresultsAUC$AUC)
longresultsAUC$AUC<-longresultsAUC$AUC/100

longresultslogloss<-gather(results, Groups, LogLoss, TrainRFlogloss, Test_RFlogloss, Test_LRlogloss, UMCUval_RFlogloss, UMCUval_LRlogloss, factor_key=TRUE)
longresultslogloss$LogLoss<-as.numeric(longresultslogloss$LogLoss)


# Visualize with ggplot
lines <- "black"
plot1<-ggplot(longresultsaccuracy, aes(x = Groups, y = Accuracy, fill=Groups))+
geom_boxplot(colour = lines, outlier.color = "black")+
scale_fill_manual(values=c("#999999", "#4DBBD5B2", "#00A087B2","#4DBBD5B2", "#00A087B2"))+
scale_color_manual(values=c("#999999", "#4DBBD5B2", "#00A087B2","#4DBBD5B2", "#00A087B2"))+
  theme(
    legend.position = "none",
    panel.border = element_blank(),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "grey"),
    legend.key=element_blank(),
    legend.background=element_blank(),
    text=element_text(size=16),
    axis.title.x=element_blank(),
    axis.text.x=element_blank())+
    xlab("Model Groups")+
    ylab("Accuracy")

plot2<-ggplot(longresultsAUC, aes(x = Groups, y = AUC, fill=Groups))+
geom_boxplot(colour = lines, outlier.color = "black")+
scale_fill_manual(values=c("#999999", "#4DBBD5B2", "#00A087B2","#4DBBD5B2", "#00A087B2"))+
scale_color_manual(values=c("#999999", "#4DBBD5B2", "#00A087B2","#4DBBD5B2", "#00A087B2"))+
  theme(
    legend.position = "none",
    panel.border = element_blank(),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "grey"),
    legend.key=element_blank(),
    legend.background=element_blank(),
    text=element_text(size=16),
    axis.title.x=element_blank(),
    axis.text.x=element_blank())+
    xlab("Model Groups")+
    ylab("AUC")

plot3<-ggplot(longresultslogloss, aes(x = Groups, y = LogLoss, fill=Groups))+
geom_boxplot(colour = lines, outlier.color = "black")+
scale_fill_manual(values=c("#999999", "#4DBBD5B2", "#00A087B2","#4DBBD5B2", "#00A087B2"))+
scale_color_manual(values=c("#999999", "#4DBBD5B2", "#00A087B2","#4DBBD5B2", "#00A087B2"))+
scale_y_continuous(labels=function(x){sprintf("%.3f", x)})+
  theme(
    legend.position = "none",
    panel.border = element_blank(),  
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "grey"),
    legend.key=element_blank(),
    legend.background=element_blank(),
    text=element_text(size=16),
    axis.text.x = element_text(face="bold", color="black", size=10, angle=45, hjust=1))+
    scale_x_discrete(labels=c("TrainRF\nOOB AUC" = "Internal\nTrain Error", "Test_RFAUC" = "RF\nTestset", "Test_LRAUC" = "RF+RR\nTestset", "UMCUval_RFAUC"= "RF\nUMCUset", "UMCUval_LRAUC" = "RF+RR\nUMCUset"))+
    xlab("Model Groups")+
    ylab("Log Loss")

```

# Plot of results

```{r echo=FALSE, fig.height = 5, fig.width = 6, fig.align = "center"}
plot_grid(plot1, plot2, plot3, align = "v", nrow = 3, rel_heights = c(1/6, 1/6, 1/2.75))
```


# Summary of results
```{r message=FALSE, warning=FALSE}
results[,c(7:21)]<-apply(results[,c(7:21)], 2, as.numeric)
summary(results[,c(7:21)])

```



# Summary of prediction scores in combined test cohorts
```{r}
predRFlist<-list()
for (r in 1:3){
  for (f in 1:3){
    tempdf<-as.data.frame(REPs[[r]][[f]][[7]][[3]])
    colnames(tempdf)<-c(paste("IlealNET",(r-1)*3+f),paste("PanNET",(r-1)*3+f),paste("PulmNET",(r-1)*3+f))
    predRFlist[[((r-1)*3)+f]]<-tempdf
  }}

predLRlist<-list()
for (r in 1:3){
  for (f in 1:3){
    tempdf<-as.data.frame(REPs[[r]][[f]][[8]][[3]])
    colnames(tempdf)<-c(paste("IlealNET",(r-1)*3+f),paste("PanNET",(r-1)*3+f),paste("PulmNET",(r-1)*3+f))
    predLRlist[[((r-1)*3)+f]]<-tempdf
    }}

predRFlist_full<-merge(predRFlist[[1]], predRFlist[[2]], by.x = "row.names", by.y = "row.names", all = TRUE)
predRFlist_full<-merge(predRFlist_full, predRFlist[[3]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predRFlist_full<-merge(predRFlist_full, predRFlist[[4]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predRFlist_full<-merge(predRFlist_full, predRFlist[[5]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predRFlist_full<-merge(predRFlist_full, predRFlist[[6]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predRFlist_full<-merge(predRFlist_full, predRFlist[[7]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predRFlist_full<-merge(predRFlist_full, predRFlist[[8]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predRFlist_full<-merge(predRFlist_full, predRFlist[[9]], by.x = "Row.names", by.y = "row.names", all = TRUE)

predLRlist_full<-merge(predLRlist[[1]], predLRlist[[2]], by.x = "row.names", by.y = "row.names", all = TRUE)
predLRlist_full<-merge(predLRlist_full, predLRlist[[3]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predLRlist_full<-merge(predLRlist_full, predLRlist[[4]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predLRlist_full<-merge(predLRlist_full, predLRlist[[5]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predLRlist_full<-merge(predLRlist_full, predLRlist[[6]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predLRlist_full<-merge(predLRlist_full, predLRlist[[7]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predLRlist_full<-merge(predLRlist_full, predLRlist[[8]], by.x = "Row.names", by.y = "row.names", all = TRUE)
predLRlist_full<-merge(predLRlist_full, predLRlist[[9]], by.x = "Row.names", by.y = "row.names", all = TRUE)

PD_combined<-read.csv("./PD_FILES/PD_combined.csv", header=TRUE)
predRFlist_full<-merge(PD_combined[,c("Sample_Name", "Sample_Group")], predRFlist_full,  by.x = "Sample_Name", by.y = "Row.names")
predLRlist_full<-merge(PD_combined[,c("Sample_Name", "Sample_Group")], predLRlist_full,  by.x = "Sample_Name", by.y = "Row.names")


predRFlist_pan<-subset(predRFlist_full, Sample_Group == "PanNET")
predRFlist_pan<-select(predRFlist_pan, matches("PanNET|Sample"))
row.names(predRFlist_pan)<-predRFlist_pan[,1]
predRFlist_pan[,1:2]<-NULL

predLRlist_pan<-subset(predLRlist_full, Sample_Group == "PanNET")
predLRlist_pan<-select(predLRlist_pan, matches("PanNET|Sample"))
row.names(predLRlist_pan)<-predLRlist_pan[,1]
predLRlist_pan[,1:2]<-NULL


predRFlist_pulm<-subset(predRFlist_full, Sample_Group == "PulmNET")
predRFlist_pulm<-select(predRFlist_pulm, matches("PulmNET|Sample"))
row.names(predRFlist_pulm)<-predRFlist_pulm[,1]
predRFlist_pulm[,1:2]<-NULL

predLRlist_pulm<-subset(predLRlist_full, Sample_Group == "PulmNET")
predLRlist_pulm<-select(predLRlist_pulm, matches("PulmNET|Sample"))
row.names(predLRlist_pulm)<-predLRlist_pulm[,1]
predLRlist_pulm[,1:2]<-NULL


predRFlist_il<-subset(predRFlist_full, Sample_Group == "IlealNET")
predRFlist_il<-select(predRFlist_il, matches("IlealNET|Sample"))
row.names(predRFlist_il)<-predRFlist_il[,1]
predRFlist_il[,1:2]<-NULL

predLRlist_il<-subset(predLRlist_full, Sample_Group == "IlealNET")
predLRlist_il<-select(predLRlist_il, matches("IlealNET|Sample"))
row.names(predLRlist_il)<-predLRlist_il[,1]
predLRlist_il[,1:2]<-NULL

predsummary <- matrix(nrow=6, ncol=7)

predsummary[1,1:6]<-summary(as.vector(as.matrix(predRFlist_il)))[1:6]
predsummary[1,7]<-sd(as.vector(as.matrix(predRFlist_il)), na.rm = TRUE)
predsummary[2,1:6]<-summary(as.vector(as.matrix(predRFlist_pan)))[1:6]
predsummary[2,7]<-sd(as.vector(as.matrix(predRFlist_pan)), na.rm = TRUE)
predsummary[3,1:6]<-summary(as.vector(as.matrix(predRFlist_pulm)))[1:6]
predsummary[3,7]<-sd(as.vector(as.matrix(predRFlist_pulm)), na.rm = TRUE)

predsummary[4,1:6]<-summary(as.vector(as.matrix(predLRlist_il)))[1:6]
predsummary[4,7]<-sd(as.vector(as.matrix(predLRlist_il)), na.rm = TRUE)
predsummary[5,1:6]<-summary(as.vector(as.matrix(predLRlist_pan)))[1:6]
predsummary[5,7]<-sd(as.vector(as.matrix(predLRlist_pan)), na.rm = TRUE)
predsummary[6,1:6]<-summary(as.vector(as.matrix(predLRlist_pulm)))[1:6]
predsummary[6,7]<-sd(as.vector(as.matrix(predLRlist_pulm)), na.rm = TRUE)

predsummary<-as.data.frame(predsummary, sort = FALSE)
row.names(predsummary)<-c("ileal RF","pancreatic RF","lung RF", "ileal LR","pancreatic LR","lung LR")
colnames(predsummary)<-c("Min.","1st Qu.","Median", "Mean","3rd Qu.","Max.", "SDev")
predsummary

myfile <- file.path(paste0("./03_nestedCV/predsummaryCV",".rData"))
save(predsummary, file = myfile)


df_il<-cbind(as.data.frame(na.omit(as.vector(as.matrix(predRFlist_il)))),as.data.frame(na.omit(as.vector(as.matrix(predLRlist_il)))))
df_pan<-cbind(as.data.frame(na.omit(as.vector(as.matrix(predRFlist_pan)))),as.data.frame(na.omit(as.vector(as.matrix(predLRlist_pan)))))
df_pulm<-cbind(as.data.frame(na.omit(as.vector(as.matrix(predRFlist_pulm)))),as.data.frame(na.omit(as.vector(as.matrix(predLRlist_pulm)))))



plot1<-ggplot(df_il, aes(x=df_il[,1], fill="red")) + 
  geom_density(alpha=0.5)+
  geom_density(data=df_il, aes(x=df_il[,2], fill="blue"), alpha=0.5)+
  xlim(0,1)+
  labs(x="Class score", y= "Density", title = "Ileal Class", fill = "Model")+
  scale_fill_identity(name = "Model fit",
                        breaks = c("red", "blue"),
                        labels = c("Raw RF", "Calibrated RF"),
                        guide = "legend")+
  theme_classic()
  
plot2<-ggplot(df_pan, aes(x=df_pan[,1], fill="red")) + 
  geom_density(alpha=0.5)+
  geom_density(data=df_pan, aes(x=df_pan[,2], fill="blue"), alpha=0.5)+
  xlim(0,1)+
  labs(x="Class score", y= "Density", title = "Pancreatic Class", fill = "Model")+
  scale_fill_identity(name = "Model fit",
                      breaks = c("red", "blue"),
                      labels = c("Raw RF", "Calibrated RF"),
                      guide = "legend")+
  theme_classic()

plot3<-ggplot(df_pulm, aes(x=df_pulm[,1], fill="red")) + 
  geom_density(alpha=0.5)+
  geom_density(data=df_pulm, aes(x=df_pulm[,2], fill="blue"), alpha=0.5)+
  xlim(0,1)+
  labs(x="Class score", y= "Density", title = "Lung Class", fill = "Model")+
  scale_fill_identity(name = "Model fit",
                      breaks = c("red", "blue"),
                      labels = c("Raw RF", "Calibrated RF"),
                      guide = "legend")+
  theme_classic()
 

plot_grid(plot1, plot2, plot3, align = "v", nrow = 3, rel_heights = c(1/3, 1/3, 1/3))

ggsave("classscorespredictiveCV.pdf", path = "./03_nestedCV/", width = 6,
       height = 4,
       units = "in")
```

