---
title: "analysis-paired-cases"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

The UMCU test cohort contains two instances of pancreatic NETs with matching data from multiple tissues. One is a pair of primary/metastasis, the other one a triplet with one primary and two metastases. 

In this analysis I used data for 65 SNPs included in the EPIC array to confirm samples come from the same patient and to select the pairs in a data-driven way. To answer the question whether the methylation profile is retained between primaries and metastases I looked at the similarity of (1) beta values and (2) RF classification output. 


Overview of matched cases: 

| Case 1                      | Case 2                     |
| --------------------------- | -------------------------- |
| 201533570026_R01C01 (Prim)  | 201533570026_R02C01 (M1)   | Patient 1, comparison 1
| 201533570026_R01C01 (Prim)  | 201533570026_R03C01 (M2)   | Patient 1, comparison 2
| 201533570026_R02C01 (M1)    | 201533570026_R03C01 (M2)   | Patient 1, comparison 3
| 202081130239_R04C01 (M)     | 202081130239_R05C01 (Prim) | Patient 2, comparison 1




## Data Import

```{r, echo = FALSE}
library(tidyverse)
library(minfi)
library(pheatmap)
library(cowplot)
library(randomForest)

# include some global helper functions
# helper function for automated preprocessing
preprocess_wrapper <- function(basenames, out_dir, preprocessFunction = NULL){
  data <- minfi::read.metharray(basenames = basenames)
  fun_norm <- ifelse(is.null(preprocessFunction),  minfi::preprocessIllumina, preprocessFunction)
  data <- fun_norm(data)
  data <- minfi::getBeta(data)
  for (i in colnames(data)){
    saveRDS(object = data[, i, drop = FALSE], file = file.path(out_dir, paste0(i, ".rds")))
  }
}

cor_mat_to_tibble <- function(mat, values){
  mat <- as_tibble(x = mat, rownames = "source")
  mat <- mat %>% 
    pivot_longer(cols = 2:ncol(.),
                 names_to = "target",
                 values_to = values,
                 values_drop_na = TRUE)
}

clean_data <- function(tib){
  
  # change sample name
  sn <- str_extract(colnames(tib)[2], pattern ="^[^_]*" )
  sn <-  paste0("il_", sn)
  
  # filter based on p-value
  pfilt <- tib[, 3, drop = TRUE] > 0.01
  tib <- tib[!pfilt, 1:2]
  
  # change column names, return
  colnames(tib) <- c("id_ref", sn)
  return(tib)
}


impute_random <- function(matrix){
  
  n_na <- sum(is.na(matrix))
  message(paste0("Imputing n = ", n_na, " missing values"))
  
  if (n_na == 0) {
    return(matrix)
  } else {
    matrix[is.na(matrix)] <- sample(x = matrix[!(is.na(matrix))], size = n_na)
    return(matrix)
  }
}


# list basenames
list_basenames <- function(dir){
  basenames <- list.files(path = dir,
                          full.names = TRUE,
                          pattern = "_Grn.idat")
  basenames <- str_replace(string = basenames, pattern = "_Grn.idat", replacement = "")
  return(basenames)
}

# upper triangle matrix
upper_tri <- function(mat){
  diag(mat) <- NA
  mat[lower.tri(mat)] <- NA
  return(mat)
}
```


```{r, results='hide', echo = FALSE}

# Sample annotation (cleaned manually beforehand)
metadata <- read_csv(file = "./PD_FILES/PD_UMCUsamples.csv", 
    skip = 7)
metadata$id<-paste0(metadata$Sentrix_ID,"_",metadata$Sentrix_Position)
metadata$Sample_Group[metadata$Sample_Group=="CasePan"]<-"PanNET"
metadata$Sample_Group[metadata$Sample_Group=="CaseIleum"]<-"IlealNET"
metadata$Sample_Group[metadata$Sample_Group=="CaseLung"]<-"PulmNET"
metadata$Sample_Group[metadata$Sample_Group=="CaseUP"]<-"PulmNET"

# Random Forest model (also, extract probe names)
load("./09_finalmodel/finalRF_and_LRseed13tr500llambda.1sea0nf3.rData")  
rf_probes <- finalRF$RFmodel$forest$xlevels %>%
  unlist %>%
  names

# Methylation Data
data_dir <- "./RAW/UMCU/IdatFiles/"
basenames <- list_basenames(dir = data_dir)

raw_data <- read.metharray(basenames = basenames, force = TRUE)
preprocessed <- minfi::preprocessIllumina(raw_data)

# Extract Beta values & SNP values
snp_values <- getSnpBeta(raw_data)
beta_values <- minfi::getBeta(preprocessed)

# Random Imputation for Beta values
set.seed(13)
beta_values <- impute_random(beta_values)
beta_values_rf <- beta_values[rf_probes, ]
#

```


## Calculate correlations
```{r, fig.height=6, fig.width=6}

# Calculate correlations
cor_beta <- cor(beta_values) %>% 
  upper_tri %>% 
  cor_mat_to_tibble(values = "cor_beta")

cor_beta_rf <- cor(beta_values_rf) %>% 
  upper_tri %>% 
  cor_mat_to_tibble(values = "cor_beta_rf")

cor_snp <- cor(snp_values) %>% 
  upper_tri %>% 
  cor_mat_to_tibble(values = "cor_snp")

pheatmap(cor(snp_values))


```

## Combine data and add annotation 

```{r}

# Merge
cor_data <- Reduce(f = full_join, x = list(cor_snp, cor_beta, cor_beta_rf))

# Extract tumor type from metadata
cor_data <- cor_data %>% 
  add_column(diag_source = metadata$Sample_Group[match(cor_data$source, metadata$id)], 
             diag_target = metadata$Sample_Group[match(cor_data$target, metadata$id)])

# Add additional variables (matched vs. ummatched cases and same vs. different NET types)
cor_data <- cor_data %>% 
  mutate(diag_group = ifelse(diag_source == diag_target, "within", "between"), 
         matched = ifelse(cor_snp > 0.8, "matched", "unmatched"))

# take a look
head(cor_data)

# look at paired samples
cor_data %>%
  filter(matched == "matched")

```


## Look at Beta Value correlations

```{r}

# all probes
cor_data %>% 
  filter(diag_source != "UpNET") %>% 
  ggplot(aes(diag_source, cor_beta)) +
  geom_boxplot(aes(fill = diag_source)) +
  facet_wrap(~ diag_group) +
  labs(x = NULL, y = "Correlation Beta Values (all probes)") +
  theme_cowplot(font_size = 12) +
  theme(legend.position = "none")
  
# only RF probes
cor_data %>% 
  filter(diag_source != "UpNET") %>% 
  ggplot(aes(diag_source, cor_beta_rf)) +
  geom_boxplot(aes(fill = diag_source)) +
  facet_wrap(~ diag_group) +
  labs(x = NULL, y = "Correlation Beta Values (Only RF probes)") +
  theme_cowplot(font_size = 12) +
  theme(legend.position = "none")
 


## correlation of matched vs. unmatched samples within PanNETs

# all probes
cor_data %>% 
  filter(diag_source == "PanNET") %>% 
  filter(diag_group == "within") %>% 
  pivot_longer(cols = c("cor_beta", "cor_beta_rf"), names_to = "probes", values_to = "cor_beta") %>% 
  mutate(probes = ifelse(probes == "cor_beta", "All Probes", "RF Probes")) %>% 
  ggplot(aes(matched, cor_beta)) +
  geom_boxplot(aes(fill = matched)) +
  facet_wrap(~ probes) +
  labs(x = NULL, y = "Correlation Beta Values") +
  theme_cowplot(font_size = 12) +
  theme(legend.position = "none")

dir.create("06_analysis_paired_cases")
ggsave("PairedCases.eps", path = "./06_analysis_paired_cases/", width = 6,
  height = 4,
  units = "in")

```


## Compare RF model output

```{r}
rf_scores_raw <- predict(object = finalRF$RFmodel, newdata = t(beta_values_rf), type = "prob")
rf_class <- predict(object = finalRF$RFmodel, newdata = t(beta_values_rf), type = "response")

rf_scores <- rf_scores_raw %>% 
  as.data.frame %>% 
  as_tibble(rownames = "id") %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "class", values_to = "class_score")

rf_data <- full_join(rf_scores, cor_data, by = c("id" = "source"))
rf_data <- full_join(rf_data, rf_scores, by = c("target" = "id"))

rf_data <- rf_data %>% 
  dplyr::rename(class_source = class.x, 
                class_score_source = class_score.x, 
                class_target = class.y, 
                class_score_target = class_score.y, 
                source = id)
rf_data <- rf_data %>% 
  relocate(source, target, diag_source, diag_target, diag_group, matched, cor_snp,
           cor_beta, cor_beta_rf, class_source, class_target, class_score_source, class_score_target)

rf_data %>% 
  filter(diag_source == diag_target) %>% 
  filter(class_source == class_target) %>%
  ggplot(aes(class_score_source, class_score_target)) +
  geom_point(fill = "steelblue", size = 3, shape = 21) +
  facet_wrap(~ matched) +
  theme_cowplot(12) +
  labs(x = "RF scores Sample 1", y = "RF scores Sample 2")

ggsave("PairedCasesRF.eps", path = "./06_analysis_paired_cases/", width = 6,
  height = 4,
  units = "in")

rf_data %>% 
  filter(diag_source == diag_target) %>% 
  filter(class_source == class_target) %>%
  group_by(matched) %>% 
  summarise(rf_score_cor = cor(class_score_source, class_score_target))
```







